\section{Conclusion}
\label{sec:conclusion}

We have presented a systematic investigation of ControlNet training configurations, focusing on the interplay between backbone freezing (\texttt{sd\_locked}) and condition injection depth (\texttt{only\_mid\_control}) under small-scale data conditions.

\subsection{Key Findings}

\begin{enumerate}
    \item \textbf{Configuration D is optimal}: Setting \texttt{sd\_locked=False} and \texttt{only\_mid\_control=True} achieves the best balance of conditional controllability, prior preservation, and generalization on the Fill50K dataset.
    
    \item \textbf{Emergent convergence is configuration-dependent}: The timing and stability of the alignment transition varies significantly, with Configuration D exhibiting stable emergence at $\sim$2400 steps.
    
    \item \textbf{Injection depth is as important as backbone freezing}: Configuration C demonstrates that multi-level injection can cause overfitting even with a frozen backbone.
    
    \item \textbf{Middle-block injection provides implicit regularization}: Restricting conditions to the semantic bottleneck constrains effective capacity while preserving adaptation flexibility.
\end{enumerate}

\subsection{Practical Recommendations}

Based on our analysis, we recommend the following strategies for training ControlNet on limited data:

\begin{enumerate}
    \item \textbf{Progressive unfreezing with smooth scheduling}: Begin with frozen backbone, gradually unfreeze layers, and use cosine annealing to smooth the alignment transition.
    
    \item \textbf{Prioritize structural layer injection}: Default to \texttt{only\_mid\_control=True} or implement hierarchical gating with middle-layer emphasis.
    
    \item \textbf{Add consistency regularization}: Consider cycle consistency or condition-extractable consistency losses:
    \begin{equation}
        \mathcal{L}_{\text{cycle}} = \| \text{Extract}(G(c)) - c \|_2^2
    \end{equation}
    
    \item \textbf{Automated hyperparameter search}: Use Bayesian optimization to explore the configuration space of unfreezing degree, injection levels, and learning rate schedules.
\end{enumerate}

\subsection{Future Directions}

Our findings open several avenues for future work:
\begin{itemize}
    \item Extending the analysis to other condition types (depth, segmentation, normal maps)
    \item Developing learnable injection gating mechanisms
    \item Investigating transfer across datasets with distribution shift
    \item Scaling analysis to larger backbone models (SDXL, SD3)
\end{itemize}

The core insight—that effective capacity and injection topology jointly shape optimization dynamics and generalization—provides a foundation for designing more robust controllable generation systems.

